Collaborative Infrastructure for Test-Driven\\Scientific Model Validation

### 2014-03-14 sciunit-icse14/sciunit-icse14.tex
ICSE14: Revised through most of sec3, added some TODOs

A \textit{SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} interface (line 1). 
A test takes one required parameter in its constructor (inherited from the base class) containing the observations against which the test will validate models.
For the class \verbx{PositionTest}, this observation parameter is a dictionary containing two types of observations (documented per Python conventions on lines 2-6). 
%For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
To create a \emph{particular} position test, we instantiate this class with particular observations. 
For example, the subset of cosmologists interested specifically in Saturn might instantiate a test by randomly chunking observations made about Saturn's nightly position over time as follows:
\vspace{-3px}
\begin{python}
  h, p = randomly_chunk(saturn_obs_positions)
  saturn_position_test = PositionTest(
    {'histories': h, 'positions': p})
\end{python}
\vspace{-5px}
The class \verb|PositionTest| defines logic that is not specific to any particular planet, so it is contained in a repository shared by all cosmologists called \verbx{cosmounit}. The particular test above would be constructed in a separate repository focused specifically on Saturn called \verbx{saturnsuite}. Both would be collaboratively developed by these (overlapping) research communities in source code repositories on Github.

@@

To produce a more comprehensive test suite, the contributors to \verbx{saturnsuite} would instantiate a number of other tests and then create an instance of the \verbx{TestSuite} class:
\begin{python}
saturn_motion_suite = sciunit.TestSuite([
  saturn_position_test, saturn_velocity_test, ...])
\end{python}
They would also instantiate a number of models. A test suite can be used to judge multiple models at once (challenge 2) if they satisfy the union of the capabilities required by the constituent tests. The result is a \emph{score matrix}:
\begin{python}
sm_matrix = saturn_motion.judge([lin_copernicus_sat, ptolemy_sat, newton_sat, kepler_sat])
\end{python}

If constructed inside an iPython notebook, a score matrix can then be visualized as an interactive table, much like the one shown in Figure \ref{fig:sciunit_overview}. Scientists can sort by column and click on tests, models and scores to get more information on each (not shown, implementation in progress).

@@

%\subsection{Models}
%Capabilities are \emph{implemented} by models. In \textit{SciUnit}, models are instances of Python classes that inherit from \verbx{sciunit.Model}. Like tests, the class itself represents a family of models, parameterized by the arguments of the constructor. A particular model is an instance of such a class.
%
%Figure \ref{fig:simple_model} shows how to write a simple {family} of models, \verbx{LinearModel}, that implement the capability in Fig. \ref{fig:capability} as well as another capability shown in Fig. \ref{fig:training}, which we will discuss below. 
%Models in this family generate a spike count by applying a linear transformation to the mean of the provided input current. The family is parameterized by the scale factor and the offset of the transformation, both scalars. 
%To create a \emph{particular} linear model, a modeler can provide particular parameter values, just as with test families:
%\begin{python}
%CA1_linear_model_heuristic = LinearModel(3.0, 1.0)
%\end{python}
%Here, the parameters to the model were picked by the modeler heuristically, or based on externally-available knowledge. 
%An alternative test design would add a training phase where these parameters were fit to data using the capability shown in Fig. \ref{fig:training}. 
%This test could thus only be used for those models for which parameters can be adjusted without human involvement. 
%Whether to build a training phase into the test protocol is a choice left to each test development community. 
%
%Fig. \ref{fig:rate_test} does not include a training phase. 
%If training data is externally available, models that nevertheless do implement a training capability (like \verb|LinearModel|) can simply be trained explicitly by calling the capability method just like any other Python method:
%\begin{python}
%CA1_linear_model_fit = LinearModel()
%CA1_linear_model_fit.train_with_currents(CA1_training_in, CA1_training_out)
%\end{python}
%\begin{figure}
%\begin{python}
%class LinearModel(sciunit.Model, SpikeCountFromCurrent, 
%    TrainSpikeCountFromCurrent):
%  def __init__(self, scale=None, offset=None): 
%    self.scale, self.offset = scale, offset
%    
%  def spike_count_from_current(self, input):
%    return int(self.scale*numpy.mean(input) + self.offset)
%
%  def train_with_currents(self, currents, counts):
%    means = [numpy.mean(c) for c in currents]
%    [self.offset, self.scale] = numpy.polyfit(means, counts, deg=1)    
%\end{python}
%\caption{A model that returns a spike count by applying a linear transformation to the mean input current. The parameters can be provided manually or learned from data provided by a test or user (see text).}
%\label{fig:simple_model}
%\vspace{-15px}
%\end{figure}
%
\section{Collaborative Workflow}\label{scidash}
A model that performs well across tests in such a suite could reasonably claim (e.g. to reviewers) to be a coherent, valid model of Saturn's motion. As new data is collected, new tests can be added to the suite or existing tests can be refined. As long as the interface between the test and the model remains the same, the table can be updated entirely automatically. Similarly, when a new model is developed, it can immediately be evaluated against all known data that has been encoded as a test by simply exposing its predictions via the capabilities the test requires. If better statistical tests are developed, they too can be used to regenerate the table automatically, as discussed above.

The design just described has been purposefully left as simple as possible, in pursuit of challenge 4. Despite its simple design, it captures the essential elements of the scientific model validation process and satisfies the four challenges we laid out. SciUnit can be used by individual scientists to organize their workflows, but because science, and in particular, model validation is a collaborative process, we have also described the intended use of SciUnit within a collaborative workflow, mediated by a social coding tool like Github.

More directly: we anticipate common testing logic (e.g. \verbx{PositionTest}), modeling logic and capabilities being collaboratively developed by larger communities in less specialized repositories like \verbx{cosmounit}. Individuals and small groups will then parameterize these tests and models with data they have gathered, as well as data known from the literature and data contained in existing data repositories to create test suites in repositories like \verbx{saturnsuite}. The interface between data collection software and existing data representation standards and tests will be mediated by bridge logic also contained in repositories like \verbx{cosmounit}.  % This workflow is meant to evoke the \emph{test-driven development} methodology that is often used in software projects.

Statisticians wishing to promote new validation metrics can simply for a $X$\verbx{unit} repository and implement new test logic. By reusing the same interfaces as previous tests used, these new metrics can immediately be used to validate or invalidate a large body of existing models, providing valuable data for use in convincing the community to adopt these into the mainstream repositories. Similarly, investigators who wish to de-emphasize or emphasize different quantities can fork $X$\verbx{suite} repository and add or remove tests.

To help organize these repositories, a simple collaborative portal sitting above Github called \textit{SciDash} is currently under development (\url{http://scidash.org/}). SciDash consists of an organized, collaboratively filtered listing of $X$\verbx{unit} and $X$\verbx{suite} repositories. That is, it is a tool to help scientists find the most popular repositories in their research area, to facilitate the development of community standards. SciDash also supports extracting summaries of tests, models, scores and related data from suite repositories to generate hyperlinked score tables and documentation. This facilitates exploratory workflows. To modify a suite (e.g. by adding the model a scientist is developing to it), a scientist can simply fork the repository. SciDash automatically discovers public forks of repositories that it is already indexing.

\section{Discussion}\label{discussion}
Academic peer review and code review are similar, but software developers typically supplement human review with test-driven methodologies. Such methodologies may benefit scientific communities as well, but several unique  constraints make reusing existing testing frameworks and processes difficult. We describe a core testing framework that addresses these issues  by using an existing, widely-adopted language and designing a flexible, simple framework that captures the domain-specific structure of scientific model validation. We then describe, by example, how the various components can be organized into software repositories and collaboratively maintained using social coding tools to support existing scientific practices. We end by outlining a collaboratively filtered portal that sits atop these tools and facilitates exploratory analyses and repository discovery.

While we discuss a toy example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in a project called NeuronUnit (http://github.com/scidash/neuronunit). We have used this as part of a collaboration with a large-scale neuroinformatics project (http://www.opensourcebrain.org), whose developers are enthusiastic about a practical pipeline for testing the dozens of models they currently maintain. Much future work remains to be done to investigate whether these tools are truly usable and useful to scientific communities, to further develop the community workflow and infrastructure and to develop a range of realistic case studies. Nevertheless, we believe that identifying the synergies between testing practices and model validation, and describing the basic tooling, represents a novel contribution to the literature on software testing. The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/. SciDash is under active development, and is currently capable of basic forms of the operations described in Sec. 3.

We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields. Typically such competitions use \emph{ad hoc} infrastructure. A common framework could be used to make designing and continuously operating such contests more feasible. 
 %In particular, we consider modeling competitions in science as a form of test-driven validation. For example, the quantitative single neuron modeling competition (QSNMC) \cite{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes, and 
%examples from other subfields of biology abound (e.g. \url{http://www.the-dream-project.org}).  These competitions typically use \emph{ad hoc} infrastructure. A common framework could be used to make designing and continuously operating such contests more feasible. In related work that

@@

\section*{Acknowledgements}
We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their helpful discussions.  
The work was supported in part by grant R01MH081905 from the National Institute of Mental Health. 
The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.

>>>>

A {SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} interface (line 1). 
A test takes one required parameter in its constructor (inherited from the base class) containing the observation(s) against which the test will validate models.
In our example, this is a dictionary documented per Python conventions on lines 2-6. 
%For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
To create a \emph{particular} position test, we instantiate this class with particular observations. 
For example, the subset of cosmologists interested specifically in Saturn might instantiate a test by randomly chunking observations made about Saturn's nightly position over time as follows:
\vspace{-3px}
\begin{python}
  h, p = randomly_chunk(saturn_obs_positions)
  saturn_position_test = PositionTest(
    {'histories': h, 'positions': p})
\end{python}
\vspace{-5px}
The class \verb|PositionTest| defines logic that is not specific to any particular planet, so it is contained in a repository shared by all cosmologists called \verbx{cosmounit}. The particular test above would be constructed in a separate repository focused specifically on Saturn called \verbx{saturnsuite}. Both would be collaboratively developed by these (overlapping) research communities in source code repositories on Github.

@@

To produce a more comprehensive test suite, the contributors to \verbx{saturnsuite} would instantiate a number of other tests and then create an instance of the \verbx{TestSuite} class:
\begin{python}
saturn_motion_suite = sciunit.TestSuite([
  saturn_position_test, saturn_velocity_test, ...])
\end{python}
They would also instantiate a number of models. A test suite can be used to judge multiple models at once (challenge 2), if they satisfy the union of the capabilities required by the constituent tests. The result is a \emph{score matrix}:
\begin{python}
sm_matrix = saturn_motion_suite.judge([lin_saturn_model, ptolemy_model, kepler_model])
\end{python}

If constructed inside an IPython notebook, a score matrix can then be visualized as an interactive table, much like the one shown in Figure \ref{fig:sciunit_overview}. Scientists can sort by column and click on tests, models and scores to get more information on each, including documentation, authorship information and related data extracted from the underlying objects.

@@

%\subsection{Models}
%Capabilities are \emph{implemented} by models. In \textit{SciUnit}, models are instances of Python classes that inherit from \verbx{sciunit.Model}. Like tests, the class itself represents a family of models, parameterized by the arguments of the constructor. A particular model is an instance of such a class.
%
%Figure \ref{fig:simple_model} shows how to write a simple {family} of models, \verbx{LinearModel}, that implement the capability in Fig. \ref{fig:capability} as well as another capability shown in Fig. \ref{fig:training}, which we will discuss below. 
%Models in this family generate a spike count by applying a linear transformation to the mean of the provided input current. The family is parameterized by the scale factor and the offset of the transformation, both scalars. 
%To create a \emph{particular} linear model, a modeler can provide particular parameter values, just as with test families:
%\begin{python}
%CA1_linear_model_heuristic = LinearModel(3.0, 1.0)
%\end{python}
%Here, the parameters to the model were picked by the modeler heuristically, or based on externally-available knowledge. 
%An alternative test design would add a training phase where these parameters were fit to data using the capability shown in Fig. \ref{fig:training}. 
%This test could thus only be used for those models for which parameters can be adjusted without human involvement. 
%Whether to build a training phase into the test protocol is a choice left to each test development community. 
%
%Fig. \ref{fig:rate_test} does not include a training phase. 
%If training data is externally available, models that nevertheless do implement a training capability (like \verb|LinearModel|) can simply be trained explicitly by calling the capability method just like any other Python method:
%\begin{python}
%CA1_linear_model_fit = LinearModel()
%CA1_linear_model_fit.train_with_currents(CA1_training_in, CA1_training_out)
%\end{python}
%\begin{figure}
%\begin{python}
%class LinearModel(sciunit.Model, SpikeCountFromCurrent, 
%    TrainSpikeCountFromCurrent):
%  def __init__(self, scale=None, offset=None): 
%    self.scale, self.offset = scale, offset
%    
%  def spike_count_from_current(self, input):
%    return int(self.scale*numpy.mean(input) + self.offset)
%
%  def train_with_currents(self, currents, counts):
%    means = [numpy.mean(c) for c in currents]
%    [self.offset, self.scale] = numpy.polyfit(means, counts, deg=1)    
%\end{python}
%\caption{A model that returns a spike count by applying a linear transformation to the mean input current. The parameters can be provided manually or learned from data provided by a test or user (see text).}
%\label{fig:simple_model}
%\vspace{-15px}
%\end{figure}
%
\section{Collaborative Workflow}\label{scidash}

Social coding tools like Github allow each scientist to work on their own fork of a repository where they can examine the consequences of their own contributions on  the tests and models that have previously been developed, facilitating both model and data sharing, and the organic adoption of community standards. For example, an experimentalist who has gathered new data can instantiate a new test or refine an existing one  with more accurate observations. As long as the interface between the test and the model remains the same, the scores can be recomputed entirely automatically, requiring no coordination with the original model developers (challenge 2). Similarly, when a modeler develops a new model, she can immediately  validate it against all the tests it is capable of taking. If a statistician develops a better goodness-of-fit metric, it too can be turned into a collection of tests that uses existing capabilities, and so the table can be regenerated automatically, as discussed above. If a scientist is not interested in certain tests, she can locally remove them to recalibrate her view of the relative merits of different models (challenge 5).

Once satisfied, these scientists might submit a \emph{pull request} to the central repository (either separately or in conjunction with the submission of a conventional paper). Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before accepting it into the ``academic record''. A model that performs well across tests in a collaboratively developed suite could more easily claim (e.g. to reviewers) to be a coherent, valid model of, for example, Saturn's motion.

%Because new models, tests and data are being added by a variety of people, there needs to be a collaborative workflow in place. As we have outlined above, the basic mechanisms have already been put in place by software developers.

To help help scientists discover these repositories and agree on which ones are central, we are developing a collaborative portal called \textit{SciDash} (\url{http://scidash.org/}). It is essentially a wiki pointing scientists to the $X$\verbx{unit}and $X$\verbx{suite} repositories designated as ``primary'' by early adopters in their research areas (who serve as \emph{de facto} initial editors as well). SciDash also supports loading IPython notebooks containing generated tables stored in these repositories into the browser directly so that score tables can be explored without going to the trouble of cloning the repository locally. To modify the suite (e.g. by adding the model a scientist is developing to it), a scientist can simply fork the repository by clicking a button at the top of such a view. We are exploring ways of loading forks directly into a cloud-based IPython notebook system, so that the complexities of Git and installation of Python and the packages required by the repository are left behind the scenes (challenge 4). SciDash discovers public forks of repositories that it is indexing and lists them to help scientists get a better sense of who is working on the problems they are interested in and what partial results are publicly available.

TODO: How to get adoption? Run contests -- dash double meaning. Also ask people directly (worked for Scholarpedia...) We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields. Typically such competitions use \emph{ad hoc} infrastructure. A common framework could be used to make designing and continuously operating such contests more feasible.

%More directly: we anticipate common testing logic (e.g. \verbx{PositionTest}), modeling logic and capabilities being collaboratively developed by larger communities in less specialized repositories like \verbx{cosmounit}. Individuals and small groups will then parameterize these tests and models with data they have gathered, as well as data known from the literature and data contained in existing data repositories to create test suites in repositories like \verbx{saturnsuite}. The interface between data collection software and existing data representation standards and tests will be mediated by bridge logic also contained in repositories like \verbx{cosmounit}.  % This workflow is meant to evoke the \emph{test-driven development} methodology that is often used in software projects. 
%
%Statisticians wishing to promote new validation metrics can simply for a $X$\verbx{unit} repository and implement new test logic. By reusing the same interfaces as previous tests used, these new metrics can immediately be used to validate or invalidate a large body of existing models, providing valuable data for use in convincing the community to adopt these into the mainstream repositories. Similarly, investigators who wish to de-emphasize or emphasize different quantities can fork $X$\verbx{suite} repository and add or remove tests.

\section{Discussion}\label{discussion}
The design just described has been purposefully left as simple as possible, in pursuit of challenge 4. Despite its simple design, it captures the essential elements of the scientific model validation process and satisfies the four challenges we laid out. TODO: SciUnit is essentially an embedded DSL that captures precisely the standard workflow of model validation. SciUnit can be used by individual scientists to organize their workflows, but because science, and in particular, model validation is a collaborative process, we have also described the intended use of SciUnit within a collaborative workflow, mediated by a social coding tool like Github.

%Academic peer review and code review are similar, but software developers typically supplement human review with test-driven methodologies. Such methodologies may benefit scientific communities as well, but several unique  constraints make reusing existing testing frameworks and processes difficult. We describe a core testing framework that addresses these issues  by using an existing, widely-adopted language and designing a flexible, simple framework that captures the domain-specific structure of scientific model validation. We then describe, by example, how the various components can be organized into software repositories and collaboratively maintained using social coding tools to support existing scientific practices. We end by outlining a collaboratively filtered portal that sits atop these tools and facilitates exploratory analyses and repository discovery.

While we discuss a toy example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in a project called NeuronUnit (http://github.com/scidash/neuronunit). We have used this as part of a collaboration with a large-scale neuroinformatics project (http://www.opensourcebrain.org), whose developers are enthusiastic about a practical pipeline for testing the dozens of models they currently maintain. TODO: More SE-research-friendly future work discussion Much future work remains to be done to investigate whether these tools are truly usable and useful to scientific communities, to further develop the community workflow and infrastructure and to develop a range of realistic case studies. Nevertheless, we believe that identifying the synergies between testing practices and model validation, and describing the basic tooling, represents a novel contribution to the literature on software testing. The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/. SciDash is under active development, and is currently capable of basic forms of the operations described in Sec. 3.

 %In particular, we consider modeling competitions in science as a form of test-driven validation. For example, the quantitative single neuron modeling competition (QSNMC) \cite{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes, and 
%examples from other subfields of biology abound (e.g. \url{http://www.the-dream-project.org}).  These competitions typically use \emph{ad hoc} infrastructure. A common framework could be used to make designing and continuously operating such contests more feasible. In related work that

@@

\section*{Acknowledgements}
We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their helpful discussions.  
The work was supported in part by NIMH grant R01MH081905. 
The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH.
